---
title: "Violas Group Presentation V1"
author: "Joe, Jason, Pam"
date: "August 4, 2018"
output: html_document
---

#Review the Guns dataset from the "AER"" package.      
* **Dataset is a balanced panel of data on 50 US states, plus the District of Columbia (for a total of 51 states), by year for 1977?????1999, and contains 1,173 observations on 13 variables. The variables are:**
    + ***state:*** factor indicating state.
    + ***year:*** factor indicating year.
    + ***violent:*** violent crime rate (incidents per 100,000 members of the population).
    + ***murder:*** murder rate (incidents per 100,000).
    + ***robbery:*** robbery rate (incidents per 100,000).
    + ***prisoners:*** incarceration rate in the state in the previous year (sentenced prisoners per 100,000 residents;value for the previous year).
    + ***afam:*** percent of state population that is African-American, ages 10 to 64.
    + ***cauc:*** percent of state population that is Caucasian, ages 10 to 64.
    + ***male:*** percent of state population that is male, ages 10 to 29.
    + ***population:*** state population, in millions of people.
    + ***income:*** real per capita personal income in the state (US dollars).
    + ***density:*** population per square mile of land area, divided by 1,000.
    + ***law:*** factor. Does the state have a shall carry law in effect in that year?


##Objective: Using the data collected see if we can Identify a model that helps best predict the violence rate.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
#install.packages("AER", repos = "http://cran.us.r-project.org", type = "source")
options(warn=-1)
library(AER)
data("Guns")
names(Guns)
```

Review a structure of the data
========================================================

```{r, echo=FALSE}
str(Guns)
```

Review data visually
========================================================
```{r, echo=FALSE, fig1, fig.height = 6, fig.width = 9, fig.align = "center"}
x <- sort(Guns$violent)
med <- median(Guns$violent)
m <- mean(Guns$violent)
std <- sd(Guns$violent)

oldpar <- par(mfrow=c(2,2))
#Price Histogram
hist(Guns$violent,main="Histogram", freq=FALSE)
curve(dnorm(x, mean=m, sd=std), col="darkblue", lwd=0.25,add = TRUE)

#Density Plot
plot(density(Guns$violent),main="Density Plot")
curve(dnorm(x, mean=m, sd=std), col="darkblue", lwd=0.25,add = TRUE)

#Sortplot
plot(sort(Guns$violent),main="Sort Plot")

#QQ-Plot
qqnorm(Guns$violent)
qqline(Guns$violent, col="blue")
par(oldpar)

```

#The violent variable does not appear to be normally distributed
* Histogram, the distribution of the violent variable deviates from the normal distribution
    + Doesn't adhere to the blue QQ-line on the the QQ-Plot. 
    + Violent variable is skewed to the right (positively) and the mean (503.074) > median (443)
* Sort Plot there appears to be a few clusters on the right side of the plot, where the violent crime rate is high
* Consider the use a ?log()? transformation on this variable possibly reduce the skewness
* Potentially make it easier to interpret whether or not a relationship exist between variables that might not have otherwise been obvious.


Scatterplot - Best way to show this?
========================================================
```{r, echo=FALSE, fig2, fig.height = 6, fig.width = 9, fig.align = "center"}
library("GGally")
ggpairs(Guns, columns = c(13, 3:11, 2), mapping = aes(color = law),lower=list(combo=wrap("facethist", binwidth=10)))
```
* The variables *murder* and *robbery* have the strongest positive linear relationship with the violent crime rate, with a correlation value of 0.827 and 0.907
+    Considering these variables are components of the violent crime rate, we expected there to be high correlation values
+    These variables are more of a subcategory of violent crimes and therefore aren't good predictors for a linear model
* Other variables that showed a high correlation value were prisoners (0.703), density (0.665), afam (0.57), cauc(-0.573), and income (0.408)
+    The law variable was also a good indicator because the box plot showed that on average states that have a shall carry law in effect have a lower violent crime rate as well as as a lower murder and robbery rate.




Fit Plot
========================================================
```{r, echo=FALSE, fig3, fig.height = 6, fig.width = 9, fig.align = "center"}
m1 <- lm(violent~., data = Guns)

library(ggplot2) #load ggplot 2 package

plot1 <- qplot(m1$fitted.value, violent, data=Guns) +
geom_abline(intercept = 0, slope = 1, color="hot pink") +
ggtitle("Fit Plot")
plot2 <- ggplot(m1, aes(.fitted, .resid)) +
geom_point() +
geom_hline(yintercept=0, color="dark blue", linetype="dashed") +
ggtitle("Residual Plot")

library(gridExtra)
grid.arrange(plot1, plot2, ncol = 2)
```
* The Fit Plot shows that the actual values fluctuate fairly significantly from the fitted values(pink reference line)
* The Residual plot shows that a majority of the observed values fall within +/-500 of the fitted values
+      This could be considered a substantial deviation from the fitted values
+      Next lets examine if a log transformation or box-cox transformation would be useful


Log Transformation Fitted Plot & Transformation Residuals vs. Fitted
========================================================
```{r, echo=FALSE, fig4, fig.height = 6, fig.width = 9, fig.align = "center"}
m2 <- lm(log(violent)~., data = Guns)
Plot1 <- qplot(m2$fitted.value, violent, data=Guns) +geom_smooth() +ggtitle("Fit Plot log transformation")

plot2 <- ggplot(m2, aes(.fitted, .resid)) +
geom_point() +
geom_hline(yintercept=0, color="dark blue", linetype="dashed") +
ggtitle("Log Transformation Residual Plot")

grid.arrange(plot1, plot2, ncol=2)
```
* The Log Transformed Fit Plot shows that the actual values fluctuate fairly significantly from the fitted values(blue reference line).
* The Transformed Fit Plot vs Residual Plot is easier to read than the previous model (m5)
+     Y-axis, which shows the .resid is shown in Z-scores with a majority of the data falling between +/- 1 standard deviations from the mean
+     Moving forward we will use the log transformation of the violent variable to make it easier to interpret the data

****Check on the m5 statement******

Box-Cox Transformation
========================================================
```{r, echo = FALSE}
library(car)
(lambda <- powerTransform(m1))
```
Should we remove this line from showing in the html?


```{r, include=FALSE}
lam <- lambda$lambda
mlam <- lm(violent^lam ~ ., Guns)
```

Box-Cox Transformation Fitted Plot & Transformation Residuals vs. Fitted
========================================================
```{r, echo = FALSE, fig5, fig.height = 6, fig.width = 9, fig.align = "center"}
plot1 <- qplot(mlam$fitted.value, violent, data=Guns) +geom_smooth() +ggtitle("Fit Plot box-cox transformation")

plot2 <- ggplot(mlam, aes(.fitted, .resid)) +
geom_point() +
geom_hline(yintercept=0, color="dark blue", linetype="dashed") +
ggtitle("box-cox Transformation Residual Plot")

grid.arrange(plot1, plot2, ncol = 2)
```

Normal QQ-plots for detecting non normality
========================================================
```{r, echo = FALSE, fig6, fig.height = 6, fig.width = 9, fig.align = "center"}
modm1 <- fortify(m1)
modm2 <- fortify(m2)
modlam <- fortify(mlam)

p1 <- qplot(sample = scale(.resid), data = modm1) + geom_abline(intercept = 0,
slope = 1, color = "red") + labs(title = "Normal QQ-Plot - Untransformed violent", y = "Residuals")

p2 <- qplot(sample = scale(.resid), data = modm2) + geom_abline(intercept = 0,
slope = 1, color = "red") + labs(title = "Normal QQ-Plot - log-Tranformed violent", y = "Residuals")

p3 <- qplot(sample = scale(.resid), data = modlam) + geom_abline(intercept = 0,
slope = 1, color = "red") + labs(title = "Normal QQ-Plot - Box-Cox", y = "Residuals")

grid.arrange(p1, p2, p3, nrow = 3)
```
Although there is not much seperation the Box-Cox Tansformed model seems to fit the best


Histogram
========================================================
```{r, echo = FALSE, fig7, fig.height = 6, fig.width = 9, fig.align = "center", message=FALSE}
p4 <- qplot(scale(.resid), data = modm1, geom = "blank") + geom_line(aes(y = ..density..,
colour = "Empirical"), stat = "density") + stat_function(fun = dnorm, aes(colour = "Normal")) + geom_histogram(aes(y = ..density..), alpha = 0.4) + scale_colour_manual(name = "Density", values = c("red", "blue")) + theme(legend.position = c(0.85, 0.85)) + labs(title = "Untransformed violent", y = "Residuals")

p5 <- qplot(scale(.resid), data = modm2, geom = "blank") + geom_line(aes(y = ..density..,
colour = "Empirical"), stat = "density") + stat_function(fun = dnorm, aes(colour = "Normal"))+ geom_histogram(aes(y = ..density..), alpha = 0.4) + scale_colour_manual(name = "Density", values = c("red", "blue")) + theme(legend.position = c(0.85, 0.85)) + labs(title = "log-Tranformed violent", y = "Residuals")

p6 <- qplot(scale(.resid), data = modlam, geom = "blank") + geom_line(aes(y = ..density..,
colour = "Empirical"), stat = "density") + stat_function(fun = dnorm, aes(colour = "Normal"))+ geom_histogram(aes(y = ..density..), alpha = 0.4) + scale_colour_manual(name = "Density", values = c("red", "blue")) + theme(legend.position = c(0.85, 0.85)) + labs(title = "box-cox Tranformed violent", y = "Residuals")
grid.arrange(p4, p5, p6, nrow = 3)
```
*The residuals of the Box-Cox Transformed model from the above plots appears to have the most normally distributed histogram


Shapiro test of Normality
========================================================
```{r, echo = FALSE}
shapiro.test(residuals(m1))
shapiro.test(residuals(m2))
shapiro.test(residuals(mlam))
```
* The Shapiro-Wilk test indicates the Pearson correlation between the residuals and the normal quantiles, W is equal to:
+     0.97023 untransformed model
+     097727 log-transformed model
+     0.98664 box-cox transformed model


**Based on our analysis we will utilize the box-cox transformed model**


Stepwise Regression
========================================================
```{r, echo=FALSE}
m3 <- step(mlam)
```
From the stepwise regression we see that cauc, density, and income both should be removed from the data to create a globally optimal model


Compare Coefficients of the two models
========================================================
```{r, echo = FALSE}
compareCoefs(mlam, m3, se = FALSE)
```
As you can see the larger model model mlam has larger standard errors then the smaller model m3, which has gone through the stepwise regression

  
P-Value of the Partial F-test
========================================================
```{r, echo = FALSE}
anova(m3, mlam)
```


*Conclusion*
========================================================
The results produced by the anova function show a non-significant result (p-value = 0.6224).  Therefore we should reject the larger model (mlam) and move forward with the smaller model (m3).



To be completed by JH and added potentially






Checking for any influential points that are controlling the results
========================================================
```{r, echo=FALSE, fig8, fig.height = 6, fig.width = 9, fig.align = "center"}
library(car)
influencePlot(mlam, id.method=cooks.distance(mlam), id.n=4)
```
* As all of the Cooks values are low then no single varaible is controlling the results




Transformation Check
========================================================
```{r, echo=FALSE, fig9, fig.height = 6, fig.width = 9, fig.align = "center"}
Guns1 <- Guns
Guns1$state <- as.numeric(Guns1$state)
Guns1$year <- as.numeric(Guns1$year)
Guns1$law <- as.numeric(Guns1$law)
Guns1$prisoners <- as.numeric(Guns1$prisoners)

stripchart(data.frame(scale(Guns1)), 
           vertical = TRUE, 
           method = "jitter")
```
* After looking at the scaled scatterplot of that data we determine that cauc needs to be transformed



Stepwise regression with cauc normalized with log
========================================================
```{r, echo = FALSE}
m4 <- lm(violent ~ year + 
           murder + 
           robbery + 
           prisoners + 
           afam + 
           log(cauc) + 
           male +
           population +
           income +
           density +
           state +
           law, 
         data = Guns)
m4 <- step(m4)
```
Interestingly this leads to the stepwise regression only from removing density and cauc from the data set but leaving income.


```{r, echo = FALSE}
anova(m3, m4, mlam)
```


```{r, include=FALSE}
library(DAAG)
```





Cross-Validation for Linear Models
========================================================
```{r, echo = FALSE, fig10, fig.height = 6, fig.width = 9, fig.align = "center"}
seed <- round(runif(1, min=0, max=100))

CVlm(data = Guns, 
     form.lm=m3, 
     m=3, 
     printit=F)
```




Cross-Validation for Linear Models against 10 seeds each
========================================================
```{r, echo = FALSE, fig11, fig.height = 6, fig.width = 9, fig.align = "center"}
df <- data.frame(mse.m1=NULL, 
                 mse.m2=NULL, 
                 mse.m3=NULL,
                 mse.m4=NULL,
                 mse.mlam=NULL)
for (i in 1:10) {
  seed <- round(runif(1, min=0, max=100))
  oldpar <- par(mfrow=c(1,5))
  mse.m1 <- CVlm(data = Guns, 
               form.lm=m1, 
               m=4, 
               seed=seed, 
               printit=F,
               main = "m1")
  mse.m2 <- CVlm(data = Guns, 
               form.lm=m2, 
               m=4, 
               seed=seed, 
               printit=F,
               main = "m2")
  mse.m3 <- CVlm(data = Guns, 
               form.lm=m3, 
               m=4, 
               seed=seed, 
               printit=F,
               main = "m3")
  mse.m4 <- CVlm(data = Guns, 
               form.lm=m4, 
               m=4, 
               seed=seed, 
               printit=F,
               main = "m4")
  mse.mlam <- CVlm(data = Guns, 
               form.lm=mlam, 
               m=4, 
               seed=seed, 
               printit=F,
               main = "mlam")
  par(oldpar)
  df.temp <- data.frame(mse.m1=attr(mse.m1, "ms"),
                        mse.m2=attr(mse.m2, "ms"),
                        mse.m3=attr(mse.m3, "ms"),
                        mse.m4=attr(mse.m4, "ms"),
                        mse.mlam=attr(mse.mlam, "ms"))
  df <- rbind(df,df.temp)
}
```

```{r, echo = FALSE}
df
```

```{r, echo = FALSE}
anova(m3, mlam)
```
```


```{r finish, include=FALSE}
options(warn = 0)
```